---
title: "lab_02"
author: "derek willis"
date: "2023-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1** Create a new column that divides the expense amounts into the following categories - \$0-\$500, \$500-\$5000, \$5001-\$50000, \$50001\_\$100000, \$100001-\$500000 - giving each an appropriate text value, and count how many expenses are in each group. How would you describe the distribution of expenses in a sentence? Would you change the categories?

**A1.** Tier 1 ($100001-$5000000) has 103 donations. Tier 2 ($50001-$100000) has 162 donations. Tier 3 ($5001-$50000) has 3589 donations. Tier 4 ($501-$5000) has 20991 donations. Tier 5 ($0-$500) has 73005 donations. There are more small-quantity donations. I'd make the categories smaller, with zero to 100 dollars, 101-1000 dollars, 1001-10000 dollars, 10001-50000 dollars, 50001-100000 dollars, 100001-1000000 dollars, and then greater than a million. 
```{r}
maryland_expenses <- maryland_expenses %>%
  mutate(
    tiered_donations = amount,
    tiered_donations = case_when(
      amount <= 500 ~ "Tier 5",
      amount >= 501 & amount <= 5000 ~ "Tier 4",
      amount >= 5001 & amount <= 50000 ~ "Tier 3",
      amount >= 50001 & amount <= 100000 ~ "Tier 2",
      amount >= 100001 & amount <= 500000 ~ "Tier 1"
    )
  )

maryland_expenses %>%
  group_by(tiered_donations) %>% 
  summarise(
    count_tiered_donations = n()
  ) 

  glimpse(maryland_expenses)

```


**Q2.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? 
**A2.** 26 departments (I counted) and Atmospheric and Oceanic science offers the most.

```{r}
open_climate_courses <- umd_courses %>% 
  filter(seats >= 1) %>% 
  filter(str_detect(title, "Climate"))

open_climate_courses %>%
  group_by(department) %>% 
  summarise(
    count_course = n()
  ) %>% 
  arrange(desc(count_course))


```

**Q3.** Of the departments listed in A2, which one is the most interesting to you, and why?
**A3.** Germanic studies because no other cultural studies have courses with "climate" in their titles.

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.
**A4.** Fixed the date format, 

```{r}
maryland_expenses <- read_csv("data/maryland_expenses.csv")
maryland_expenses <- maryland_expenses %>% mutate(expenditure_date=mdy(expenditure_date))


head(maryland_expenses)
```

**Q5.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.
**A5.** 

```{r}
```
