---
title: "Class Reference"
author: "Sarah Meklir"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Howdy!")
```



## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/Documents/GitHub/data_journalism_2023_spring")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### How to read a CSV file

The function to use here is called "read_csv" and we pass it either a file path or a URL.

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```
## How to count rows

You use the 'n()' function to count rows, like so:
```{r}
dataframe %>% 
  group_by(some_column) %>% 
  summarize(count = n())
```

## How to write a filter
```{r}
journalism_courses <- umd_courses %>% filter(department == "Journalism")
```
so name it, then carrot arrow, the data frame, pipe character, filter(operator == "case sensitive filter thing")

## How to find certain things in data
```{r}
umd_courses %>% 
  filter(str_detect(title, "Shakespeare"))
```
very cool

## make sure date stuff is chill:
maryland_expenses <- maryland_expenses %>% mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses)

then sort/group by month and find amounts
maryland_expenses %>%
  mutate(month = floor_date(expenditure_date, "month")) %>%
  group_by(month) %>%
  summarise(total_amount = sum(amount)) %>%
  arrange(desc(total_amount))
  
## combining multiple charts with the same data but different years into one (bind_rows with list)
county_voters_combined <- bind_rows(list(county_voters_2016, county_voters_2018, county_voters_2020, county_voters_2022))
View(county_voters_combined)

## to combine population chart with voter number chart with nrow included to show row total
maryland_population %>% left_join(county_voters_2022, by="COUNTY") %>% nrow()

## to combine pop chart w voter pop chart
  maryland_population_with_voters <- maryland_population %>% left_join(county_voters_2022)
  
## combining things when the column names don't align
maryland_population_with_voters_and_demographics <- maryland_population_with_voters %>% left_join(maryland_demographics, by=c("COUNTY"="NAME"))

## how to change a datatype to character
maryland_zcta <- maryland_zcta %>% mutate(ZCTA5N = as.character(ZCTA5N))
glimpse(maryland_zcta)
