---
title: "lab_05"
author: "Derek Willis"
date: "2023-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). Read the questions carefully and attempt to do all of what they ask. Make a list if it helps.

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(janitor)
library(lubridate)
```

For this lab, we want to investigate spending by Maryland state and local candidates during the 2022 election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis. The function to export data from RStudio is called `write_csv()` and it takes two arguments: the name of the dataframe and the path of the file to write it to, which is in quotes. Like `read_csv()`, but the variable name comes inside. Here's an example, in which you'd replace the dataframe and the path values.

```{r}
write_csv(dataframe, "folder/file.csv")
```


## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table, clean the column names and convert columns with dates into actual dates.
maryland_expenses <- read_csv("~/Documents/data_journalism_2023_spring/labs/lab_04/data/maryland_expenses.csv")
  cleaned_maryland_expenses <- maryland_expenses %>%
    clean_names() %>% 
    mutate(expenditure_date=mdy(expenditure_date))
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". Write code to isolate Media expenses where the name of the payee is not NA. Then, using the `write_csv` function, make a CSV file of those media expenses in your data folder. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing your work on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: https://about.fb.com/news/2021/10/facebook-company-is-now-meta/).

Then, when you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving that file a different name* and read it back into RStudio as a new dataframe.

Then, using that `payee_clean` column, write code to find all of the spending where Facebook is the payee and calculate the total amount of spending for each committee and present the results showing the committee that spent the most first. Which committee is that?

**A1.** Okay! It looks like the first committee to spend money on Facebook was "Jain  Ashwani Friends of". I think what you're asking is which committee spent the most money in the shortest span of time, and I'm not sure how to find that. I can, however, tell you that the committee that spent the most money on Facebook overall was "WFP National Pac - Non-Contribution Account," so. Go them, I guess. Although that also might not be true, because I'm pretty sure I didn't combine the committees with the same names.

```{r}
cleaned_maryland_expenses <- cleaned_maryland_expenses %>% 
  filter(str_detect(expense_category, "Media")) %>% 
  filter(payee_name != "NA") 

payee_clean <- cleaned_maryland_expenses %>% 
  filter(str_detect(payee_name, "FACEBOOK") | str_detect(payee_name, "META")) %>% 
  group_by(committee_name, expenditure_date) %>% 
  summarize(total_spending = sum(amount)) %>% 
  arrange(desc(total_spending)) 
  
write_csv(payee_clean, "~/Documents/data_journalism_2023_spring/labs/lab_04/data/payee_clean.csv")
## So I did this to create the payee-clean.csv, but then forgot to change workspaces after so it's also partly a draft for my second part of this first question. I'm not sure what it looked like before, but it definitely worked for what I needed it to do before I overwrote it with what is also below.
```
```{r}
refined_payee_clean <- read_csv("~/Documents/data_journalism_2023_spring/labs/lab_04/data/payee-clean-csv.csv") %>% 
  filter(str_detect(payee_name, "FACEBOOK") | str_detect(payee_name, "META")) %>% 
  group_by(committee_name, payee_name, expenditure_date) %>% 
  summarize(total_spending = sum(amount)) %>% 
  arrange(desc(total_spending)) 

refined_payee_clean
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library (https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\[direction\]=desc&sort_data\[mode\]=relevancy_monthly_grouped&media_type=all) and look for ads from the top spender. What do you notice about those ads?

**A2.** 
1. WFP National Pac - Non-Contribution Account: It's a national PAC, they stand for working families.
2. Franchot Anderson-Walker for Maryland Slate: Franchot and Anderson-Walker were a ticket running for Governor
3. Casa in Action PAC: a group endorsing democratic candidates in elections. In MD, they backed Wes Moore. They support the voting rights and activity of Latinos and people of color.
4. Jain Ashwani Friends of: Ashwani Jain was running for governor
5. Ashburn James Friends to Elect: James Ashburn was running for State Delegate.
6, 7. Casa in Action PAC: this means I didn't combine the combined totals for committees' expenditures. 
8. Franchot Peter Friends Of: He was a candidate for Governor in 2022
9. Franchot Anderson-Walker for Maryland Slate: Another repeat! Dang.
10. Fani Gonzalez Natali for Council: Candidate for Montgomery County Council


**Q3.** Given the list of top spenders in A1 and the primary results for governor (https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html), who is missing from that A1 list? Did those campaigns run Facebook ads, according to Facebook's ad library? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees)?

**A3.** Dan Cox wasn't on there, or Wes Moore. But they both ran ads according to the Facebook ad library. The biggest recipients were "SKDK" and "FOGLAMP" which seems... unlikely.

```{r}
wes_moore_payments_maybe <- read_csv("~/Documents/data_journalism_2023_spring/labs/lab_04/data/payee-clean-csv.csv") %>% 
  filter(str_detect(committee_name, "Moore  Wes For Maryland")) %>% 
  group_by(payee_name) %>% 
  summarize(total_spending = sum(amount)) %>% 
  arrange(desc(total_spending)) 

wes_moore_payments_maybe
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** There's more variety in this list, and it looks like the only repeat in the top ten is Casa in Action PAC. And it looks like they spent a whole lot more on online advertising in general than on Facebook. I'm a little surprised Calvin Ball's team spent so much on online advertising and I live in the area and still can't remember seeing ads for him online.

```{r}
online_advertising_data <- read_csv("~/Documents/data_journalism_2023_spring/labs/lab_04/data/payee-clean-csv.csv") %>% 
  filter(str_detect(expense_purpose, "Online Advertising")) %>% 
  group_by(committee_name) %>% 
  summarize(total_spending = sum(amount)) %>% 
  arrange(desc(total_spending)) 

online_advertising_data
```

**Q5** You've been asked to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** I think my best story idea would be from the spread of democratic versus republican candidates' spending and where they targeted that money. Did more democrats advertise online? Where? Did republicans focus on TV ads or physical posters? It would be interesting to explore. As for takeaways, I know a lot of money was spent, but I don't have a great idea of how much was spent on facebook because I wasn't looking for the right things in my code.
